---
title: "Analyzing US Births and Basketball Recruits"
author: "Pushpendu Biswas"
format: html
execute:
  echo: false
---

```{r}
#| message: false
library(tidyverse)
library(readxl)
```

```{r}
read_excel("data/us_births_1994_2014.xlsx") |>
  glimpse()
```

```{r}
read_excel("data/us_births_1994_2014.xlsx") |>
  select(births, year) |>
  summary()
```

```{r}
#| cache: true
births_tibble <- read_excel("data/us_births_1994_2014.xlsx") |>
  mutate(day_of_week = factor(day_of_week,
                              levels = c("Sun", "Mon", "Tues", "Wed", "Thurs", "Fri", "Sat"),
                              ordered = TRUE))
```

```{r}
births_tibble |>
  group_by(month, date_of_month) |>
  summarize(avg_births = mean(births), .groups = "drop")
```

```{r}
births_tibble |>
  group_by(month, date_of_month) |>
  summarize(avg_births = mean(births), .groups = "drop") |>
  ggplot(aes(x = date_of_month, y = month, fill = avg_births)) +
  geom_tile() +
  scale_y_reverse(breaks = 1:12, labels = month.name) +
  scale_fill_gradient(low = "lightgreen", high = "darkgreen") +
  theme_minimal() +
  labs(
    x = "Date of Month",
    y = "Month",
    fill = "Mean Births"
  )
```

## December 25 Births Comparison

The following analysis computes, for each year, the number of births on December 25 and compares it to the average number of births on the surrounding days (December 20–24 and 27–30). The December 25 births are then expressed as a percentage of that baseline.

```{r}
dec25_births <- births_tibble |>
  filter(month == 12, date_of_month == 25) |>
  select(year, births_dec25 = births)

surrounding_births <- births_tibble |>
  filter(month == 12, date_of_month %in% c(20:24, 27:30)) |>
  group_by(year) |>
  summarize(births_surrounding_avg = mean(births), .groups = "drop")

christmas_data <- dec25_births |>
  left_join(surrounding_births, by = "year") |>
  mutate(pct_of_baseline = (births_dec25 / births_surrounding_avg) * 100)

# Table display
christmas_data |>
  knitr::kable(
    col.names = c("Year", "Births on Dec 25", "Avg Births (Dec 20–24 & 27–30)", "Percentage of Baseline (%)"),
    digits = 1,
    caption = "December 25 Births vs. Surrounding Days Baseline (1994–2014)"
  )
```

We can summarize this dataset using the `summary()` function:

```{r}
christmas_data |>
  summary()
```

We can also visualize the percentage of baseline over time using a line and point chart:

```{r}
christmas_data |>
  left_join(
    births_tibble |>
      filter(month == 12, date_of_month == 25) |>
      select(year, day_of_week),
    by = "year"
  ) |>
  ggplot(aes(x = year, y = pct_of_baseline)) +
  geom_line(color = "darkred", linewidth = 1) +
  geom_point(aes(color = day_of_week), size = 2) +
  scale_x_continuous(breaks = seq(1994, 2014, 2)) +
  labs(
    title = "December 25 Births as a Percentage of Surrounding Days Baseline",
    x = "Year",
    y = "Percentage of Baseline (%)",
    color = "Day of Week"
  ) +
  theme_minimal()
```

## Modeling Births

We fit a linear model to predict the number of births using year, month, and day of the week:

```{r}
births_model <- lm(births ~ factor(year) + factor(month) + day_of_week, data = births_tibble)
summary(births_model)$r.squared
```

```{r}
births_adjusted <- births_tibble |>
  mutate(pct_resid = (residuals(births_model) / mean(births)) * 100)
glimpse(births_adjusted)
```

```{r}
calendar_resid <- births_adjusted |>
  filter(!(month == 2 & date_of_month == 29)) |>
  group_by(month, date_of_month) |>
  summarize(mean_pct_resid = mean(pct_resid), .groups = "drop") |>
  mutate(calendar_date = as.Date(paste(2001, month, date_of_month, sep = "-")))
```

We can now plot `calendar_resid` to visualize the average percentage deviation by calendar date:

```{r}
calendar_resid |>
  ggplot(aes(x = calendar_date, y = mean_pct_resid)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey40") +
  annotate("text", x = as.Date("2001-01-01"), y = -24.3, label = "New Year's Day", hjust = -0.1, color = "darkred", size = 3) +
  annotate("text", x = as.Date("2001-07-04"), y = -26.1, label = "Independence Day", hjust = -0.1, color = "darkred", size = 3) +
  annotate("text", x = as.Date("2001-12-25"), y = -39.4, label = "Christmas Day", hjust = 1.1, color = "darkred", size = 3) +
  scale_x_date(date_labels = "%b", date_breaks = "1 month") +
  labs(
    title = "Average Birth Residuals by Day of the Year",
    subtitle = "Percentage deviation from regression baseline, 1994-2014",
    x = "Date (2001 Anchor Year)",
    y = "Mean Residuals (%)"
  ) +
  theme_minimal()
```

## NBA Recruits Data Analysis

We load the NBA recruits dataset to analyze player statistics and ratings:

```{r}
nba_recruits <- read_excel("data/nba_recruits.xlsx")
nba_recruits
```

```{r}
read_excel("data/nba_recruits.xlsx") |>
  glimpse()
```

```{r}
read_excel("data/nba_recruits.xlsx") |>
  select(rank, nba_mean_ws48, top_mean_wa, total_seasons, drafted) |>
  summary()
```

```{r}
read_excel("data/nba_recruits.xlsx") |>
  count(tier)
```

```{r}
read_excel("data/nba_recruits.xlsx") |>
  mutate(tier = factor(tier,
                       levels = c("Never played", "Brief career", "Solid career", "All-Star level", "Superstar"), 
                       ordered = TRUE)) |>
  mutate(recruit_group = factor(recruit_group,
                                levels = c("#1–10", "#11–25", "#26–50", "#51–100", "Outside top 100"),
                                ordered = TRUE)) |>
                                count(recruit_group)
```

```{r}
#| cache: true
basketball_tibble <- read_excel("data/nba_recruits.xlsx") |>
  mutate(tier = factor(tier,
                       levels = c("Never played", "Brief career", "Solid career", "All-Star level", "Superstar"), 
                       ordered = TRUE)) |>
  mutate(recruit_group = factor(recruit_group,
                                levels = c("#1–10", "#11–25", "#26–50", "#51–100", "Outside top 100"),
                                ordered = TRUE))
basketball_tibble |> glimpse()                     
```

```{r}
basketball_tibble |>
  filter(!is.na(rank) & !is.na(top_mean_wa)) |>
  ggplot(aes(x = rank, y = top_mean_wa)) +
  geom_point(aes(color = tier), alpha = 0.8, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  geom_text(
    data = function(df) filter(df, name %in% c("LeBron James", "Kevin Durant", "Kawhi Leonard", "Draymond Green")),
    aes(label = name),
    vjust = -0.7,
    hjust = 0.5,
    size = 3
  ) +
  scale_color_viridis_d() +
  labs(
    title = "NBA Recruit Rank vs. Top Mean Win Average",
    subtitle = "Showing a linear trend across career tiers",
    x = "Recruit Rank",
    y = "Top Mean Win Average (top_mean_wa)",
    color = "Career Tier"
  ) +
  theme_minimal()
```
